Extended Early Adulthood and Attitude Change

### 2024-06-25 manuscript/formative_period_kk.Rmd
kkedits624

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Early adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage. We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame. We find strikingly little difference in the rates of opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates of change. 
---

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change. Panel duration varied from less than one year (the 1980 ANES Major Panel Study) to five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

@@

Model 1 in Table XXX presents the main effects and variance components of the equation presented above.

The intercept term in Model 1 suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration, on any particular question. This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.67. Many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.

Second, duration between observations is positively but weakly associated with wave-to-wave absolute change. For every additional year between waves, respondents' absolute deviance increases about .71 points. Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14. Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].

The main effect of year of observation in decades is negative and also small (-.37). On average, respondents exhibit less wave-to-wave change in their responses in more recent panels. However, question-level random effects vary considerably around this main effect, with a standard deviation of about .88. This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.

>>>>

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Early adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage. We leverage these changes to explore whether changes in the structure of the life course have affected how likely people of different cohorts are to change their stated opinions over time. We find strikingly little difference in the rates of opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates of change. 
---

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experiences more shocks that should shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

@@

Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above.

The intercept term in Model 1 suggests that, on average, survey respondents are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question. This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.86. On some questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.

Second, duration between observations is positively but weakly associated with wave-to-wave absolute change. For every additional year between waves, respondents' absolute deviance increases about .71 points. Notably, variance around this main effect is relatively small (s.d. = 0.43). In other words, while some items demonstrate a stronger effect of duration than others, these differences are of comparable magnitude. Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].

Turning now to the coefficients for age groups, members of the youngest age group (18-25) are notably less stable than respondents in other age groups in the most recent panels. Respondents in the 26-33 age group are about 1.5 points more stable on average, respondents in the 34-65 age group are about 3 points more stable on average, and respondents in the 65 and over age group are about 4 points more stable on average, than respondents in the 18 to 25 age group. This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.

The main effect of year of observation in decades is negative and also small. On average, respondents exhibit less wave-to-wave change in their responses in more recent panels. However, question-level random effects vary considerably around this main effect, with a standard deviation of about .88. This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.

Turning now to the central question of whether the amount of change we observe for each group has changed over time, we turn to the interaction effects between time and age group. The coefficient for the reference category (18-25) suggests that respondents in this age group have not become any more or less stable opinion responders since the start of the observation window.
